---
title: "Using `RTCGA.PANCAN12` package to compare time to death for selected tumor types"
author: "Przemyslaw Biecek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using RTCGA.PANCAN12 to compare time to death}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(comment="", message=FALSE, warning = FALSE, tidy.opts=list(keep.blank.line=TRUE, width.cutoff=150),options(width=150), cache=TRUE)
```

# RTCGA.cnv package

You need RTCGA.PANCAN12 package to use PANCAN12 data from Cancer Genome Browser.

"/Users/pbiecek/_TCGA_wszystkie_raki_/data/genome_cancer"

```{r}
require(devtools)
if (!require(RTCGA.PANCAN12)) {
    install_github("mi2-warsaw/RTCGA.data", subdir = "RTCGA.PANCAN12")
    require(RTCGA.PANCAN12)
}
```

# Expression

Expression data is divided int to datasets, since it's too big to fit GitHub limits.
So first you need to bind both data sets into one.

```{r}
expression.cb <- rbind(expression.cb1, expression.cb2)
```

Now, let's see where is expression for MDM2 and let's create a dataset with single column single  - MDM2 expression.

```{r}
grep(expression.cb[,1], pattern="MDM2")

MDM2 <- expression.cb[8467,-1]
MDM2v <- t(MDM2)
```

# Mutations

Mutations are stored in mutation.cb and they are coded 0/1.
Let's gather all mutations for TP53 and then let's create a data frame with single column - only for TP53?

```{r}
grep(mutation.cb[,1], pattern="TP53$", value = FALSE)

TP53 <- mutation.cb[18475,-1]
TP53v <- t(TP53)
```

# Merging

From clinical data we are going to extract following variables:
- X_cohort (cancer type)
- X_EVENT (0/1)
- X_TIME_TO_EVENT (in days)
- X_PANCAN_UNC_RNAseq_PANCAN_K16 (cancer's subtypes)

And then we merge clinical, expression and mutation datasets.
Not that it requires some data cleaning (in clinical data _ is used as a separator while in expression and mutation it's .).

```{r}
dfC <- data.frame(names=gsub(clinical.cb[,1], pattern="-", replacement="."), clinical.cb[,c("X_cohort","X_TIME_TO_EVENT","X_EVENT","X_PANCAN_UNC_RNAseq_PANCAN_K16")])
dfT <- data.frame(names=rownames(TP53v), vT = TP53v)
dfM <- data.frame(names=rownames(MDM2v), vM = MDM2v)
dfTMC <- merge(merge(dfT, dfM), dfC)
colnames(dfTMC) = c("names", "TP53", "MDM2", "cohort","TIME_TO_EVENT","EVENT","PANCAN_UNC_RNAseq_PANCAN_K16")
dfTMC$TP53 <- factor(dfTMC$TP53)

# only primary tumor
dfTMC <- dfTMC[grep(dfTMC$names, pattern="01$"),]
```

# MDM2 and TP53

Firs't let's see the MDM2 expression along cancers 

```{r}
library(ggplot2)

ggplot(dfTMC, aes(x=cohort, y=MDM2)) + geom_boxplot() + theme_bw() + coord_flip() + ylab("")

```

And let's see the fraction of TP53 in cancers

```{r}
ggplot(dfTMC, aes(x=cohort, fill=TP53)) + geom_bar() + theme_bw() + coord_flip() + ylab("")
```

And how many cases for particular cancer types?

```{r}
sort(table(dfTMC$cohort))
```

# Survival in different cancer types given MDM2 and TP53

Let's dychotomize MDM2 into two groups with the cutoff =0 (almost median).

```{r}
dfTMC$MDM2b <- cut(dfTMC$MDM2, c(-100,0,100), labels=c("low", "high"))
```

Number of cases

```{r}
library(dplyr)
library(tidyr)
dfTMC %>% 
  group_by(MDM2b, TP53, cohort) %>%
  summarize(count=n()) %>%
  unite(TP53_MDM2, TP53, MDM2b) %>%
  spread(TP53_MDM2, count, fill = 0)
```

For four cancers with largest number of cases let's see Kaplan Meier curves divided into MDM2/TP53 groups.

```{r}
library(survey)
library(scales)
library(survMisc)

# cancer = "TCGA Breast Cancer"
for (cancer in names(sort(-table(dfTMC$cohort)))) {
  survp <- survfit(Surv(TIME_TO_EVENT/356,EVENT)~TP53+MDM2b, data=dfTMC, subset=cohort == cancer)
  autoplot(survp, title = "")$plot + theme_bw() + scale_x_continuous(limits=c(0,10), breaks=0:10) + ggtitle(cancer) + scale_y_continuous(labels = percent, limits=c(0,1))
}

```
