---
title: RTCGA.data - The Family Of R Packages Containing TCGA Data
author:
  - name: Marcin Kosinski
    affiliation: Warsaw University of Technology
    address:
    - Faculty of Mathematics and Information Science
    - Koszykowa 75, 00-662 Warsaw, Poland
    email:  M.P.Kosinski@gmail.com
  - name: Przemysław Biecek
    affiliation: Warsaw University
    address:
    - line 1
    - line 2
    email:  Przemyslaw.Biecek@gmail.com
abstract: >
  The following article presents RTCGA.data: a family of R packages containing The Cancer Genome Atlas Project (TCGA) data. The Cancer Genome Atlas (TCGA) is a comprehensive and coordinated effort to accelerate our understanding of the molecular basis of cancer through the application of genome analysis technologies, including large-scale genome sequencing (1). We provide TCGA data in few separate packages that are hosted on one GitHub repository, what made those luxurious data easier to possess and manage. We hope providing researchers with comprehensive catalogs of the key genomic changes in many major types and subtypes of cancer will support advances in developing more effective ways to diagnose, treat and prevent cancer. 
preamble: >
  % Any extra latex you need in the preamble
  \usepackage{wrapfig,booktabs}
output: rticles::rjournal_article
---

## Motivation

The Cancer Genome Atlas (TCGA) Data Portal provides a platform for researchers to search, download, and analyze data sets generated by TCGA. It contains clinical information, genomic characterization data, and high level sequence analysis of the tumor genomes (1). The key is to understand genomics to improve cancer care.

## Data origin

Data from TCGA are available through Firehose Broad GDAC portal (2). One can select cohort (assigned to the cancer type) and data type, i.e. clinical data, and download a `tar.gz` file with compressed data. 

The main disadvantages of such data download methodology are:

- The downloaded files are compressed `tar.gz` files and not everyone manages to unpack such files.
- If one requires to download datasets containing i.e. information about genes' expressions for all available cohorts types (TCGA collected data for more than 30 various cancer types) one would have to go through click-to-download process many times, which is inconvienent and time-consuming.
- Clinical datasets from TCGA project are not in a standard tidy data format, which is: one row for one observation and one column for one variable. They are transposed what makes work with those data burdensome. That becomes more onerous when one would like to investigate
many clinical datasets.
- Datasets containing information on gene's mutations are not in one easy-to-handle file. Every patient has it's own file, what for many potential researchers may be a impassable barrier. Just think about BRCA cohort (breast cancer) with more that 1000 various patients and more than 1200 files with patients' genes mutations information.


## RTCGA.data family data

For reasons described in previous section we prepared selected datasets from TCGA project in an easy to handle and process way and embed them in 5 separate R packages. All packages can be installed from GitHub by evaluating the following code:

```{r, eval=FALSE}
if (!require(devtools)) {
    install.packages("devtools")
    require(devtools)
}
install_github("mi2-warsaw/RTCGA.data", 
                subdir = paste0("RTCGA.", 
                                c("clinical", "rnaseq", "mutations", "cnv", "PANCAN12")
                               )
 )
```

One package, i.e. `RTCGA.clinical` can be installed with the command

```{r, eval=FALSE}
if (!require(devtools)) {
    install.packages("devtools")
    require(devtools)
}
install_github("mi2-warsaw/RTCGA.data", 
                subdir = "RTCGA.clinical") 
```

If you are using Windows, make sure you have rtools [3] installed on your computer, before evaluating aboved commands.

RTCGA.data family contains 5 packages:

- `RTCGA.clinical` package containing clinical datasets from TCGA. Each cohort contains one dataset prepared in a tidy format. Each row, marked with patients' barcode, corresponds to one patient.
- `RTCGA.rnaseq` package containing genes' expressions datasets from TCGA. Each cohort contains one dataset with over 20 thousand of columns corresponding to genes' expression. Rows correspond to patients, that can be matched with patient's barcode.
- `RTCGA.mutations` package containint genes' mutations datsets from TCGA. Each cohort contains one dataset with extra column specifying patient's barcode which enables to distinguish which rows correspond to which patient.
- `RTCGA.cnv` package \textcolor{red}{ explanation needed.}
- `RTCGA.PANCAN12` package \textcolor{red}{ explanation needed.}

More detailed information about datasets included in RTCGA.data family are shown in Table \ref{data_details}

\tiny
\begin{table}[h!]

\begin{tabular}{lrrrrrr}
\toprule
Disease Name & Cohort & *.clinical & *.mutations & *.rnaseq & *.cnv & *.PANCAN12 \\
\toprule
Adrenocortical carcinoma &	ACC & \ & \ & \ & \ & \	\\
Bladder urothelial carcinoma &	BLCA& \ & \ & \ & \ & \	\\
Breast invasive carcinoma &	BRCA	& \ & \ & \ & \ & \	\\
Cervical and endocervical cancers &	CESC	& \ & \ & \ & \ & \	\\
Cholangiocarcinoma &	CHOL	& \ & \ & \ & \ & \	\\
Colon adenocarcinoma	 & COAD	& \ & \ & \ & \ & \	\\
Colorectal adenocarcinoma &	COADREAD	& \ & \ & \ & \ & \	\\
Lymphoid Neoplasm Diffuse &	DLBC	& \ & \ & \ & \ & \	\\
Esophageal carcinoma &	ESCA	& \ & \ & \ & \ & \	\\
FFPE Pilot Phase II	 & FPPP	& \ & \ & \ & \ & \	\\
Glioblastoma multiforme	& GBM & \ & \ & \ & \ & \	\\
Glioma &	GBMLGG	& \ & \ & \ & \ & \	\\
Head and Neck squamous cell carcinoma &	HNSC& \ & \ & \ & \ & \	\\
Kidney Chromophobe &	KICH	& \ & \ & \ & \ & \	\\
Pan-kidney cohort (KICH+KIRC+KIRP) &	KIPAN	& \ & \ & \ & \ & \	\\
Kidney renal clear cell carcinoma &	KIRC	& \ & \ & \ & \ & \	\\
Kidney renal papillary cell carcinoma &	KIRP & \ & \ & \ & \ & \	\\
Acute Myeloid Leukemia &	LAML	& \ & \ & \ & \ & \	\\
Brain Lower Grade Glioma &	LGG & \ & \ & \ & \ & \	\\
Liver hepatocellular carcinoma &	LIHC & \ & \ & \ & \ & \	\\
Lung adenocarcinoma &	LUAD	& \ & \ & \ & \ & \	\\
Lung squamous cell carcinoma &	LUSC & \ & \ & \ & \ & \	\\
Mesothelioma &	MESO	& \ & \ & \ & \ & \	\\
Ovarian serous cystadenocarcinoma &	OV & \ & \ & \ & \ & \	\\
Pancreatic adenocarcinoma &	PAAD	& \ & \ & \ & \ & \	\\
Pheochromocytoma and Paraganglioma &	PCPG & \ & \ & \ & \ & \	\\
Prostate adenocarcinoma &	PRAD & \ & \ & \ & \ & \	\\
Rectum adenocarcinoma &	READ & \ & \ & \ & \ & \	\\
Sarcoma &	SARC	& \ & \ & \ & \ & \	\\
Skin Cutaneous Melanoma &	SKCM	& \ & \ & \ & \ & \	\\
Stomach adenocarcinoma &	STAD	& \ & \ & \ & \ & \	\\
Stomach and Esophageal carcinoma &	STES	& \ & \ & \ & \ & \	\\
Testicular Germ Cell Tumors &	TGCT	& \ & \ & \ & \ & \	\\
Thyroid carcinoma &	THCA	& \ & \ & \ & \ & \	\\
Thymoma &	THYM	& \ & \ & \ & \ & \	\\
Uterine Corpus Endometrial Carcinoma &	UCEC & \ & \ & \ & \ & \	\\
Uterine Carcinosarcoma &	UCS & \ & \ & \ & \ & \	\\
Uveal Melanoma &	UVM & \ & \ & \ & \ & \	\\
\bottomrule
\end{tabular}

\end{table}
\normalsize

## How to work with RTCGA.data family


## Patient's barcode as a key to merge data 


## Applications examples


```{r, eval=FALSE}
[1] http://cancergenome.nih.gov/

[2] http://gdac.broadinstitute.org/
   
[3] http://cran.r-project.org/bin/windows/Rtools/

\bibliography{RJreferences}
   
```

