---
title: "Using `RTCGA` package to download mutations data that are included in `RTCGA.mutations` package"
author: "Marcin Kosiński"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using RTCGA to download mutations data as included in RTCGA.data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(comment="", message=FALSE, warning = FALSE, tidy.opts=list(keep.blank.line=TRUE, width.cutoff=150),options(width=150), eval=FALSE)
```

# RTCGA package

> The Cancer Genome Atlas (TCGA) Data Portal provides a platform for researchers to search, download, and analyze data sets generated by TCGA. It contains clinical information, genomic characterization data, and high level sequence analysis of the tumor genomes. The key is to understand genomics to improve cancer care.

`RTCGA` package offers download and integration of the variety and volume of TCGA data using patient barcode key, what enables easier data possession. This may have an benefcial infuence on impact on development of science and improvement of patients' treatment. `RTCGA` is an open-source R package, available to download from Bioconductor 

```{r, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("RTCGA")
```

or from github
```{r, eval=FALSE}
if (!require(devtools)) {
    install.packages("devtools")
    require(devtools)
}
install_github("MarcinKosinski/RTCGA")
```

Furthermore, `RTCGA` package transforms TCGA data to form which is convenient to use in R statistical package. Those data transformations can be a part of statistical analysis pipeline which can be more reproducible with `RTCGA`.

# How to download clinical data to gain the same datasets as in RTCGA.mutations package?

There are many available date times of TCGA data releases. To see them all just type:
```{r, eval=FALSE}
library(RTCGA)
checkTCGA('Dates')
```

Version 1.0 of `RTCGA.mutations` package contains mutations datasets from `2015-06-01`.
There were downloaded as follows (which is mainly copied from [http://marcinkosinski.github.io/RTCGA/](http://marcinkosinski.github.io/RTCGA/):

## Available cohorts

All cohort names can be checked using:
```{r, eval=FALSE}
(cohorts <- infoTCGA() %>% 
   names() %>% 
   sub("-counts", "", x=.))
```

For all cohorts the following code downloads the mutations data.

## Downloading tarred files
```{r, eval=FALSE}
#dir.create( "data2" )
date <- "2015-06-01"
sapply( cohorts, function(element){
tryCatch({
downloadTCGA( cancerTypes = element, 
              dataSet = "Mutation_Packager_Calls.Level",
              destDir = "data2/", 
              date = date )},
error = function(cond){
   cat("Error: Maybe there weren't mutations data for ", element, " cancer.\n")
}
)
})
```

## Untarring downloaded files

```{r, eval=FALSE}
list.files( "data2/") %>% 
   paste0( "data2/", .) %>%
   grep( "tar.gz", x = ., value=TRUE) %>%
   sapply( untar, exdir = "data2/" )
```
## Removing no longer needed tar.gz files
After datasets are untarred, the `tar.gz` files ar no longer needed and can be deleted.

```{r, eval=FALSE}
list.files( "data2/") %>% 
   paste0( "data2/", .) %>%
   grep( pattern = "tar.gz", x = ., value = TRUE) %>%
   sapply( file.remove )

```


# Saving mutations data to `RTCGA.mutations` package


```{r, eval=FALSE}
#check this next time
list.files("data2/") %>%
   grep( x=., pattern ="mutations", value = TRUE) %>%
   paste0("data2/", .) %>%
   sapply( function(element){
      readTCGA(element,"mutations") -> mutations_file
      assign( x = substr(element, start=7, stop=100),
               value = mutations_file,
              envir = .GlobalEnv )

      save( list= substr(element, start=7, stop=100),
            file = paste0("data/",
                     substr(element, start=7, stop=100),
                     ".rda"))
      rm(list = substr(element, start=7, stop=100), envir = .GlobalEnv)
   })
```

