---
title: "R package containing TCGA data"
author: "Marcin Kosiński"
date: "`r Sys.Date()`"
output: 
   rmarkdown::html_vignette:
      fig_width: 17
      fig_height: 10
      toc: true
      toc_depth: 3
      keep_md: true
      includes:
         in_header: include/in_header.html
         before_body: include/before_body.html
vignette: >
  %\VignetteIndexEntry{home}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(comment="", message=FALSE, warning = FALSE, tidy.opts=list(keep.blank.line=TRUE, width.cutoff=150),options(width=150), cache=TRUE, eval = FALSE)
```

# RTCGA package

> The Cancer Genome Atlas (TCGA) Data Portal provides a platform for researchers to search, download, and analyze data sets generated by TCGA. It contains clinical information, genomic characterization data, and high level sequence analysis of the tumor genomes[1]. The key is to understand genomics to improve cancer care.

`RTCGA` package offers download and integration of the variety and volume of TCGA data using patient barcode key, what enables easier data possession. This may have an benefcial infuence on impact on development of science and improvement of patients' treatment. `RTCGA` is an open-source R package, available to download from Bioconductor (not there yet) or from github
```{r}
if (!require(devtools)) {
    install.packages("devtools")
    require(devtools)
}
install_github("MarcinKosinski/RTCGA")
```

Furthermore, `RTCGA` package transforms TCGA data to form which is convenient to use in R statistical package. Those data transformations can be a part of statistical analysis pipeline which can be more reproducible with `RTCGA`.

Make sure you have [rtools](http://cran.r-project.org/bin/windows/Rtools/) installed on your computer.

# RTCGA.data package

> Datasets from TCGA project

<h5> Installation of the `RTCGA.data` package: </h5>
Install the latest version of **RTCGA.data** from Bioconductor:
```{r}
# not there yet
```
or use:
```{r}
if (!require(devtools)) {
    install.packages("devtools")
    require(devtools)
}
install_github("mi2-warsaw/RTCGA.data")
```

<h5> The list of available datasets: </h5>
```{r}
library(RTCGA.data)
?clinical
?rnaseq
```
